Evaluating FAIR Digital Object and Linked Data as distributed object systems

### 2023-11-15 latex/main.tex
Typo/grammar fixes

The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been recognised by the research community and policy makers as a goal to strive for \autocite{h2020fair2016}. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}. The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

@@

These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data. Given their importance, in this article, we aim to compare FAIR Digital Objects with Linked Data and the Web architecture in the context of the discourse around FAIR data.

@@

@@

The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}. This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced by \textcite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the \textcite{x1255FrameworkDiscovery} framework which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}.

Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see \emph{Next steps for FDO} \vpageref{sec:next-step-fdo}).

@@

The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review. As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:

See bibliography \vref*{sec:fdo-bibliography} for the citation per document above. It is worth pointing out that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations. Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.

@@

\section*{Method}

@@

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).

@@

\section*{Results}

@@

Based on the analysis shown in Table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:

@@

From this mapping\footnote{An equivalent SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology. Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

@@

In Table \vref{tbl:fdo-checks} we evaluate completeness of the guidelines in two current FDO realizations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.

A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines. Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals. The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}. Likewise, a Linked Data approach will need to follow the same requirements to actually comply as an FDO implementation.

From our evaluation, we can observe:

@@

@@

From this evaluation we observe:

@@

\section*{Conclusion}\label{conclusion}

@@

@@

\printbibliography

\clearpage
\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
\printshorthands[heading=fdobibliography]

>>>>

The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been recognised by the research community and policy makers as a goal to strive for \autocite{h2020fair2016}. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}. The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of digital objects that are extensively described. Such descriptions for interoperability connect a range of information -- from protocols and presentations, to hardware designs and scientific workflows, including extensive metadata of the information itself.

@@

These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data. Given their importance, in this article, we aim to compare FDO with Linked Data and the Web architecture in the context of the discourse around FAIR data.

@@

The rest of this paper is organised as follows: We begin with a background primer on FDO and Linked Data to provide a foundation for the work. In the Methods section, we introduce the conceptual frameworks we use for comparison. Subsequently, in the Results section, we systematically step through the outcomes of applying these frameworks to both FDO and Linked Data. For each framework, we derive key observations. We end with a discussion of these results and their implications for both approaches and conclude.

@@

The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as a way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}. This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced by \textcite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the \textcite{x1255FrameworkDiscovery} framework which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}. As mentioned previously, this ecoystem is envisioned to consist of a wide variety of digital entities and contextual information ranging from software to articles to even descriptions of experimental infrastructures \autocite{Azeroual2022PuttingFP}. Recently, it has been noted that the practical use of FDOs to achieve interoperability requires governance in particular with respect to assessing such interoperability \autocite{Wilkinson2023}.

FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents.

@@

The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.  We provide an overview of these documents in Appendix \ref{sec:appendixfdo}. These documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}.  Except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations. Going forward a key strategy of the Forum is the use of profiles to help define specific attributes in metadata that are necessary for domains or application contexts. However, these are not yet fully implemented in the implementations considered here.

Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.

@@

\section*{Methods}\label{sec:methods}

@@

Conceptual framework 1, 3, 5 consider more general views of interoperability between systems, whereas frameworks 2 and 4 are developed specifically for addressing FAIR principles.

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community.

@@

For all our comparisons, our process was to perform a mapping between the relevant specifications and/or implementation and the given conceptual model through detailed reading of the defining documents. We aim in all cases for traceability between the given specification and our mapping such that readers can validate our analysis.

\section*{Results}\label{sec:results}

@@

\subsubsection*{Observations}
Based on the analysis shown in Table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:

@@

From this mapping\footnote{A SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology. Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

@@

In Table \vref{tbl:fdo-checks}, we evaluate completeness of the guidelines in two current FDO realisations: 1) DOIPv2 \autocite{foundationDigitalObjectInterface} and 2) Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}. We provide our analysis of each realization with respect to the FDO Guidline and also provide suggestions for that realization to meet the given guideline.

A key insight from this is that simply using DOIP does not achieve many of the FDO guidelines. Rather the guidelines set out how a protocol like DOIP should be used to achieve FAIR Digital Object goals. The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}. Likewise, a Linked Data approach will need to follow the same FDO requirements to actually comply as an FDO implementation.

\subsubsection*{Observations}

@@

\subsubsection*{Observations}

@@

\subsubsection*{Observations}

@@

\section*{Conclusion}\label{sec:conclusion}

@@

Views and opinions expressed in this work are those of the authors only and do not necessarily reflect those of the funded projects, FAIR Digital Object Forum, European Union nor the European Commission.

@@

\printbibliography
\clearpage

\appendix
\section{An overview of upcoming FDO specifications} \label{sec:appendixfdo}

See bibliography below for the citation per document above.

%\clearpage
\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
\printshorthands[heading=fdobibliography]
